---
title: "Project_1"
author: "Mahmoud Badr"
date: "5/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Data
```{r load, echo = TRUE }
# Load data
if (!file.exists("activity.csv") )
    {
     dlurl <- 'http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'  
     download.file(dlurl,destfile='repdata%2Fdata%2Factivity.zip',mode='wb')  
     unzip('repdata%2Fdata%2Factivity.zip')
    }

# Read data
data <- read.csv("activity.csv")  
```

## What is mean total number of steps taken per day?
* Calculate the total number of steps taken per day
* If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
* Calculate and report the mean and median of the total number of steps taken per day

```{r q1, echo = TRUE }
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="green",xlab="Number of Steps")
rmean <- mean(steps_by_day$steps)
rmean
rmedian <- median(steps_by_day$steps)
rmedian
```
The mean is `r rmean` and the median is `r rmedian`

## What is the average daily activity pattern?
* Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r q2, echo = TRUE }
steps_by_interval <- aggregate(steps ~ interval, data, mean)
plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")
max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
max_interval
```
The interval with most steps is `r max_interval`

## Imputing missing values
### Calculate and report the total number of missing values in the dataset
```{r q3, echo = TRUE }
NATotal <- sum(!complete.cases(data))
NATotal
```
Total Number of Missing values are `r NATotal`

### Using Mean for the day compute missing values
```{r q4, echo = TRUE}
StepsAverage <- aggregate(steps ~ interval, data = data, FUN = mean)
fillNA <- numeric()
for (i in 1:nrow(data)) {
    obs <- data[i, ]
    if (is.na(obs$steps)) {
        steps <- subset(StepsAverage, interval == obs$interval)$steps
    } else {
        steps <- obs$steps
    }
    fillNA <- c(fillNA, steps)
}
```

### Create a new dataset including the imputed missing values
```{r q5, echo = TRUE}
new_activity <- data
new_activity$steps <- fillNA
```

### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
``` {r q6, echo = TRUE}
StepsTotalUnion <- aggregate(steps ~ date, data = new_activity, sum, na.rm = TRUE)
hist(StepsTotalUnion$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
#Create Histogram to show difference. 
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="green", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "green"), lwd=10)
```

### Calculate Mean
```{r q7, echo = TRUE}
rmeantotal <- mean(StepsTotalUnion$steps)
rmeantotal
```

### Calculate Median
```{r q8, echo = TRUE}
rmediantotal <- median(StepsTotalUnion$steps)
rmediantotal
```

### Do these values differ from the estimates from the first part of the assignment?
```{r q9, echo = TRUE}
rmediandiff <- rmediantotal - rmedian
rmediandiff
rmeandiff <- rmeantotal - rmean
rmeandiff
```
The mean(Mean Var: `r rmeandiff`) is the same however the median does have a small variance(Median Var: `r rmediandiff`). between the total which includes the missing values to the base

* What is the impact of imputing missing data on the estimates of the total daily number of steps?

*On observation the impact of the missing data has the biggest effect on the 10000 - 150000 step interval and changes frequency from 27.5 to 35 a variance of 7.5*

## Are there differences in activity patterns between weekdays and weekends?
* Created a plot to compare and contrast number of steps between the week and weekend. There is a higher peak earlier on weekdays, and more overall activity on weekends.

``` {r q10, echo = TRUE}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
new_activity$dow = as.factor(ifelse(is.element(weekdays(as.Date(new_activity$date)),weekdays), "Weekday", "Weekend"))
StepsTotalUnion <- aggregate(steps ~ interval + dow, new_activity, mean)
library(lattice)
xyplot(StepsTotalUnion$steps ~ StepsTotalUnion$interval|StepsTotalUnion$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```